---
title: "Notes on NFL Survival Betting Strategies: An OR Perspective"
format: html
editor: visual
author: "Tim Anderson"

---

# NFL Survival Betting Strategies-An OR Perspective

## Introduction

Survival betting pools in the NFL are based on each week being in a pool where you select a team to win. If your team wins, you advance to the next week. The goal is to be the last one "surviving".

The goal of this analysis is to examine different betting strategies. There is a wide variety of selection strategies. These are generally naive algorithms.

-   Selecting the strongest team each week. Selecting the team with the strongest odds of winning each week.

-   Point spread, win probability, ELO differential, or other metric).

-   Anti-team - pick a bad team and whenever possible, pick their opponent.  For example, pick against the New York Jets.  

Factors that might affect:

-   Contrarian strategies trying to pick "unpopular" teams early (likely more important with large pools)

-   Limiting based on home field advantage (don't waste a team's week when they are on the road)

-   Remaining strength of schedule (pick a team this week before their schedule gets harder)

-   Pick teams early that are at risk of injury to key contributors (teams with high value but injury prone QB or RB should generally be picked earlier)

-   Save choices of teams that are playing against teams that might be jockeying for high draft picks for late in the season. (Tanking for draft picks is less of an issue in NBA than NFL but may still be a useful factor.)


## Datasets and Packages

There are several potential packages of data that could be useful.

nflseedR is a popular package that allows simulating NFL seasons.

### Using nflseedR

```{r}
library(nflseedR)    # NFL data and simulation
library(dplyr)       # Data munging package
library(DT)          # Data table package 
library(kableExtra)  # Also for doing tables
```

To be consistent with the vignette for the package, let's explore the 2022 season.

```{r}
all_games <- nflseedR::sims_games_example |> 
  dplyr::filter(game_type == "REG")
DT::datatable(all_games)
```

This is a listing of all the 2022 regular season games.

This does a simulation of all of the 2022 regular season games with four replications. It is chunked into two groups for allowing parallelization if the system supports it.

```{r}
sim <- nflseedR::nfl_simulations(
  games = all_games,            # All regular season games from 2022-23
  simulations = 4,              # Run 4 simulated seasons
  chunks = 2                    # Parallelizes into 2 chunks for speed 
                                #    Divides the 4 sims into 2 sims each
)
```

Let's take a step back focus on the week by week information.

With that, let's start by reviewing the schedule data set better.

```{r}
wk1_games <- nflseedR::sims_games_example |> 
  dplyr::filter(game_type == "REG")       |>
  dplyr::filter(week == 1)                
DT::datatable(wk1_games)
```

```{r}
kbl(wk1_games)
```

-   Let's look over the core schedule data in the data table.\
    First column is just a number for the game.

-   season is season, or in this case the 2022-23 season.

-   `game_type` is regular (REG) or playoff. For this analysis we only care about regular season games.

-   `week` is the week number. This has changed over the years to now 18 weeks for each team to play a 17 game schedule.

-   `away_team` and `home_team` are the 3 letter codes of the franchise/team for the away and home teams.

-   `away_rest` and `home_rest` describe the number of days of rest since their previous games for the away and home teams respectively. *Note: It defaults to 7 for the start of the season. I don't know if it adds time for teams coming off a bye or that played a Thursday night game the previous week.*

-   `location` says `Home` for each game. Presumably international or neutral site location games are coded differently.

-   result indicates the actual outcome of the game. For example, the Atlanta Falcons (`ATL`) lost their 2022 home opener to the New Orleans Saints (`NO`) 26-27 and therefore the result is -`1.`It is unclear why several games are listed as result of `NA` though as all those games were played. Perhaps this is a fluke in the sample dataset. For example, the `BUF LA` game's score was 31-10 (11) and was a Thursday game played 3 days before most of the other games.

This is sufficient for simulation to occur.

Now, let's simply simulate week 1 once.

```{r}
sim1 <- nflseedR::nfl_simulations(
  games = wk1_games,    # Week 1 regular season games from 2022-23
  simulations = 1,      # Run 1 simulated season
  chunks=1,             # Run as all in one chunk
  verbosity="NONE",     # Default is "MIN"
  tiebreaker_depth = "SOS")  # 1 week of data is not enough for tie breakers
                             # leaving at default does not work for just 1
                             # week.

```

This is odd. It worked to simulate the whole season but it appears to be

Other questions,

where do I find initial team ratings of some form?

Looks like perhaps I need to use the routine that runs inside of the simulation, `nflseedR_compute_results`.

Let's try running this case.

```{r}
#standings <- nflseedR::nfl_standings(wk1_games, ranks = "DRAFT")
#season_data <- [season == 2022]
#teams <- season_data$team

xteams <- list(sim$standings$team[1:32])
```

```{r error=TRUE}
com1 <- nflseedR::nflseedR_compute_results(
  teams = xteams,
  games = wk1_games,    # Week 1 regular season games from 2022-23
  simulations = 1,      # Run 1 simulated season
  chunks=1,             # Run as all in one chunk
  verbosity="NONE",     # Default is "MIN"
  tiebreaker_depth = "SOS")  # 1 week of data is not enough for tie breakers
                             # leaving at default does not work for just 1
                             # week.

```

## Another try at analysis

```{r}
games <- nflreadr::load_schedules(2024) |> 
  dplyr::filter(game_type == "REG") |> 
  dplyr::mutate(
    result = NA_integer_,
    away_score = NA_integer_,
    home_score = NA_integer_
  )
```

```{r }
team_ratings <- tibble::tribble(
  ~team, ~spread,
  "ARI", -2.4,
  "ATL",  0.0,
  "BAL",  3.9,
  "BUF",  2.1,
  "CAR", -5.0,
  "CHI", -0.2,
  "CIN",  2.9,
  "CLE",  1.1,
  "DAL",  2.1,
  "DEN", -4.0,
  "DET",  2.6,
  "GB",   1.8,
  "HOU",  1.8,
  "IND", -0.6,
  "JAX",  0.1,
  "KC",   4.5,
  "LA",   0.9,
  "LAC", -0.7,
  "LV",  -2.7,
  "MIA",  1.6,
  "MIN", -2.1,
  "NE",  -4.5,
  "NO",  -2.2,
  "NYG", -3.5,
  "NYJ",  2.1,
  "PHI",  2.0,
  "PIT", -0.6,
  "SEA", -1.5,
  "SF",   5.2,
  "TB",  -1.8,
  "TEN", -3.2,
  "WAS", -3.2
)
setNames(team_ratings$spread, team_ratings$team)
```

```{r}
team_ratings$elo <- 1500 + 25 * team_ratings$spread
  # Simple model for converting spread to ELO - used in package Vignette 
```

```{r}
sims <- nflseedR::nfl_simulations(
  games = games,
  simulations = 4,
  chunks = 2
)
```

### Reviewing results

Let's look over the results of this simulation. Some items: It only looks at the regular season.

-   I want look over the revised ELO ratings.

-   Need to find where they are tucked away.

```{r}
DT::datatable(sims$games, 
              caption="Simulated game results for 2024 season")

DT::datatable(sims$overall, 
              caption="Simulated overall for 2024 season")

DT::datatable(sims$game_summary, 
              caption="Simulated game summaries for 2024 season")

```

#### End of Season Results

Notice that the results show for each simulation what the result is, including the playoffs.  This is not really needed for our purpose. In our case, it is 32x4=128 simulation results (rows). Note that it will grow in size quickly with a lot of simulation results.  

```{r}
DT::datatable(sims$standings, 
              caption="Simulated end of season standings for 2024 season")
```



